---
title: 'Explore and analyze data with R'
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: breezedark
    toc: yes
    toc_float: yes
    code_download: yes
---

## **Flights Data Exploration Challenge**

A significant part of a a data scientist's role is to explore, analyze, and visualize data. In this challenge, you'll explore a real-world dataset containing flights data from the US Department of Transportation.

Let's start by loading the Tidyverse.

```{r}
# Load the packages in the tidyverse into the current R session
suppressPackageStartupMessages({
  library(tidyverse)
  })


```

Now, we can import the into R and start doing some data science on it!

```{r}
# Load and view the data
df_flights <- read_csv("https://raw.githubusercontent.com/MicrosoftDocs/ml-basics/master/challenges/data/flights.csv", show_col_types = FALSE)

df_flights %>% 
  slice_head(n = 7)
```

The dataset contains observations of US domestic flights in 2013, and consists of the following fields:

-   **Year**: The year of the flight (all records are from 2013)

-   **Month**: The month of the flight

-   **DayofMonth**: The day of the month on which the flight departed

-   **DayOfWeek**: The day of the week on which the flight departed - from 1 (Monday) to 7 (Sunday)

-   **Carrier**: The two-letter abbreviation for the airline.

-   **OriginAirportID**: A unique numeric identifier for the departure aiport

-   **OriginAirportName**: The full name of the departure airport

-   **OriginCity**: The departure airport city

-   **OriginState**: The departure airport state

-   **DestAirportID**: A unique numeric identifier for the destination aiport

-   **DestAirportName**: The full name of the destination airport

-   **DestCity**: The destination airport city

-   **DestState**: The destination airport state

-   **CRSDepTime**: The scheduled departure time

-   **DepDelay**: The number of minutes departure was delayed (flight that left ahead of schedule have a negative value)

-   **DelDelay15**: A binary indicator that departure was delayed by more than 15 minutes (and therefore considered "late")

-   **CRSArrTime**: The scheduled arrival time

-   **ArrDelay**: The number of minutes arrival was delayed (flight that arrived ahead of schedule have a negative value)

-   **ArrDelay15**: A binary indicator that arrival was delayed by more than 15 minutes (and therefore considered "late")

-   **Cancelled**: A binary indicator that the flight was cancelled

Your challenge is to explore the flight data to analyze possible factors that affect delays in departure or arrival of a flight.

1.  Start by cleaning the data.

    -   Identify any null or missing data, and impute appropriate replacement values.

    -   Identify and eliminate any outliers in the **DepDelay** and **ArrDelay** columns.

2.  Explore the cleaned data.

    -   View summary statistics for the numeric fields in the dataset.

    -   Determine the distribution of the **DepDelay** and **ArrDelay** columns.

    -   Use statistics, aggregate functions, and visualizations to answer the following questions:

        -   *What are the average (mean) departure and arrival delays?*

        -   *How do the carriers compare in terms of arrival delay performance?*

        -   *Is there a noticable difference in arrival delays for different days of the week?*

        -   *Which departure airport has the highest average departure delay?*

        -   *Do **late** departures tend to result in longer arrival delays than on-time departures?*

        -   *Which route (from origin airport to destination airport) has the most **late** arrivals?*

        -   *Which route has the highest average arrival delay?*

Add markdown and code cells as required to create your solution.

> **Note**: There is no single "correct" solution. A sample solution is provided in 01 - Flights Challenge.ipynb.

```{r}
# Your adventure starts here
```

```{r}
library(here)
library(rmd2jupyter)
rmd2jupyter("01_Data_Exploration_Challenge.Rmd")
```
