---
title: 'Train and Evaluate Regression Models using Tidymodels'
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: breezedark
    toc: yes
    toc_float: yes
    code_download: yes
---

## **Regression Challenge**

Predicting the selling price of a residential property depends on a number of factors, including the property age, availability of local amenities, and location.

In this challenge, you will use a dataset of real estate sales transactions to predict the price-per-unit of a property based on its features. The price-per-unit in this data is based on a unit measurement of 3.3 square meters.

> **Citation**: The data used in this exercise originates from the following study:
>
> *Yeh, I. C., & Hsu, T. K. (2018). Building real estate valuation models with comparative approach through case-based reasoning. Applied Soft Computing, 65, 260-271.*
>
> It was obtained from the UCI dataset repository (Dua, D. and Graff, C. (2019). [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml). Irvine, CA: University of California, School of Information and Computer Science).

## **Review the data**

Let's hit the ground running by importing the data and viewing the first few rows.

```{r}
# Load the core tidyverse and tidymodels in your current R session
suppressPackageStartupMessages({
  library(tidyverse)
  library(tidymodels)
})

# Read the csv file into a tibble
data <- read_csv(file = "https://raw.githubusercontent.com/MicrosoftDocs/ml-basics/master/challenges/data/real_estate.csv", show_col_types = FALSE)

# Print the first 10 rows of the data
data %>% 
  slice_head(n = 10)
```

The data consists of the following variables:

-   **transaction_date** - the transaction date (for example, 2013.250=2013 March, 2013.500=2013 June, etc.)

-   **house_age** - the house age (in years)

-   **transit_distance** - the distance to the nearest light rail station (in meters)

-   **local_convenience_stores** - the number of convenience stores within walking distance

-   **latitude** - the geographic coordinate, latitude

-   **longitude** - the geographic coordinate, longitude

-   **price_per_unit** house price of unit area (3.3 square meters)

## **Train a Regression Model**

Your challenge is to explore and prepare the data, identify predictive features that will help predict the **price_per_unit** label, and train a regression model that achieves the lowest Root Mean Square Error (RMSE) you can achieve (which must be less than **7**) when evaluated against a test subset of data.

Add markdown and code cells as required to create your solution.

> **Note**: There is no single "correct" solution. A sample solution is provided in [02 - Real Estate Regression Solution.ipynb](https://render.githubusercontent.com/view/02%20-%20Real%20Estate%20Regression%20Solution.ipynb).

Add markdown and code cells as required to to explore the data, train a model, and evaluate the model's predictive performance.

```{r}
# Your adventure to evaluate data, and train and evaluate a regression model begins here
```

## **Use the Trained Model**

Save your trained model, and then use it to predict the price-per-unit for the following real estate transactions:

| **transaction_date** | **house_age** | **transit_distance** | **local_convenience_stores** | **latitude** | **longitude** |
|---------------------|----------------|--------------|--------|-------|------|
| 2013.167             | 16.2          | 289.3248             | 5                            | 24.98203     | 121.54348     |
| 2013.000             | 13.6          | 4082.015             | 0                            | 24.94155     | 121.50381     |

```{r}
# Your code to use the trained model
```

```{r include=FALSE}
library(here)
library(rmd2jupyter)
rmd2jupyter("02_Regression_Challenge.Rmd")
```
